Biology background

### 2014-12-08 Rmarkdown/biology_background.Rmd
Typos and fixes
Included more figures in biology_background

One way to explore the distribution of CpGs in the human genome is to look at the distribution of distances from one CpG to the next, the _intra-pair distances_ (IPDs). __FIGURE__ is the empirical cumulative distribution function of CpG IPDs for the human reference genome (hg19). We see that approximately $70\%$ of CpGs are within 100 bp of the next CpG. __FIGURE__ also shows the expected IPD distribution under a model where CpGs are uniformly distributed along the genome with probability equal to the observed frequency of CpGs on each chromosome. By comparing the observed IPDs to the expected IPDs we see that the distribution of distances between CpGs has more "close" pairs and more "distant" pairs than we would expect by chance.

__FIGURE OBSERVED VS EXPECTED__ also shows that there is a cluster of CpGs with $IPD < 10$. These largely correspond to CpGs that lie within what are called _CpG islands_ (CGIs). Within these CpG islands are the 20-40% of CpGs that are frequently unmethylated in mammalian genomes. CpG islands are important regulatory elements in the genome and are where most differences in DNA methylation between different cell types are found \citep{Wu:2010do}. The classical definition of a CpG island, given by \citep{GardinerGarden:1987we}, and used by, for example, the UCSC genome browser, is a region of the genome where:

@@

These "sandy/beachy" metaphors have been continued (i.e., stretched to breaking point) with various authors defining CpG island shores, CpG island shelves, CpGs in the "open sea", CpG deserts and CpG canyons. CpG island shores, shelves and the open sea are all defined with respect to CpG islands. Specifically,

@@

__TODO: Describe CHARM \citep{Irizarry:2008hg}, although not yet sure which sub-section it should appear in__.

@@

The Illumina Infinium HumanMethylation450 BeadChip (_Illumina 450k_) provides a modern implementation of this type of assay for studying human DNA. This array assays $482,422$ cytosines, $99.3\%$ of these CpGs, across a wide variety of genomic features \citep{Stirzaker:2014cu}. There are well-established methods and software for analysing Illumina 450k data, which were recently reviewed by \citey{Dedeurwaerder:2014bt}.

>>>>

One way to explore the distribution of CpGs in the human genome is to look at the distribution of distances from one CpG to the next, the _intra-pair distances_ (IPDs). Figure \ref{fig:IPD_ECDF_hg19} is a plot of the empirical cumulative distribution function of CpG IPDs for the human reference genome (hg19). We see that approximately $70\%$ of CpGs are within 100 bp of the next CpG. Figure \ref{fig:IPD_ECDF_hg19} also shows the expected IPD distribution under a model where CpGs are uniformly distributed along the genome with probability equal to the observed frequency of CpGs on each chromosome. By comparing the observed IPDs to the expected IPDs we see that the distribution of distances between CpGs has more 'close' pairs and more 'distant' pairs than we would expect by chance.

\begin{figure}[htbp]
\includegraphics[width=\textwidth]{../figures/IPD_ECDF_hg19.pdf}
\caption{Plot of the empirical cumulative distribution function of distance between adjacent CpGs in the human reference genome (hg19). The observed distances are contrasts with those under the 'expected' model whereby CpGs are distributed uniformly at random with probability equal to the observed frequency of Cpgs on each chromosome.}
\label{fig:IPD_ECDF_hg19}
\end{figure}

Figure \ref{fig:IPD_frequency_polygon_hg19} is an alternative way to visualise these data, which plots the percentage of pairs of CpGs with a given IPD. Figure \ref{fig:IPD_frequency_polygon_hg19} shows that there is a cluster of CpGs with $IPD < 10$. These largely correspond to CpGs that lie within what are called _CpG islands_ (CGIs).

\begin{figure}[htbp]
\includegraphics[width=\textwidth]{../figures/IPD_frequency_polygon_hg19.pdf}
\caption{The frequency of the observed distances between adjacent CpGs in the human reference genome (hg19).}
\label{fig:IPD_frequency_polygon_hg19}
\end{figure}

CpG islands contain the 20-40% of CpGs that are frequently unmethylated in mammalian genomes. CpG islands are important regulatory elements in the genome and are where most differences in DNA methylation between different cell types are found \citep{Wu:2010do}. The classical definition of a CpG island, given by \citep{GardinerGarden:1987we}, and used by, for example, the UCSC genome browser, is a region of the genome where:

@@

These 'sandy/beachy' metaphors have been continued (i.e., stretched to breaking point) with various authors defining CpG island shores, CpG island shelves, CpGs in the "open sea", CpG deserts and CpG canyons. CpG island shores, shelves and the open sea are all defined with respect to CpG islands. Specifically,

@@

@@

The Illumina Infinium HumanMethylation450 BeadChip (_Illumina 450k_) provides a modern implementation of this type of assay for studying human DNA. This array assays $482,422$ cytosines, $99.3\%$ of these CpGs, across a wide variety of genomic features \citep{Stirzaker:2014cu}. There are well-established methods and software for analysing Illumina 450k data, which were recently reviewed by \citet{Dedeurwaerder:2014bt}.
